Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-10-21 pot/LaTeX/skewnormal.tex
time series tweaks

\subsection{Accounting for temporal dependence} \label{s:temporal}
In a standard block-maxima approach, the temporal dependence is assumed to be negligible because blocks (e.g. yearly maxima) are taken to be large enough to assume independence.
However, when using daily meausrements, the assumption of temporal independence is no longer appropriate.
We account for temporal dependence through an AR(1) time series on the spatial knots, $z_t(\bs)$ terms, and $\sigma^2_t(\bs)$ terms.
We first, transform the spatial knots from $\calD$ to $\calR^2$ as follows.
Let
\begin{align*}
  \bw^*_k = \Phi^{-1}_2\left[ \frac{ (\bw_k - \min(\bs))}{ \text{range}(\bs) } \right].
\end{align*}
where $\Phi_2$ represents a bivariate standard normal density function.
Then $\bw^*_k \in \calR^2$.
We use a copula on the $\sigma^2_t(\bs)$ to ensure that the marginal distributions of $\sigma^2_t(\bs)$ are inverse gamma.
Let
\begin{align*}
  \sigma^{2*}_t(\bs) =\Phi^{-1}\left\{ \text{IG}[\sigma^2_t(\bs)] \right\}
\end{align*}
where $\Phi$ is a univariate standard normal density function, and IG is the distribution function for an IG$(a, b)$ random variable.
Then the time series is modeled as
\begin{align}
  \bw^*_{t, k} | \bw^*_{t-1, k} &\sim N_2\left[\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2) \right] \\
  z_{t, k} | z_{t-1, k} &\sim N \left[\phi_z z_{t-1, k}, \sigma^2_{t,k} (1 - \phi_z^2)\right] \\
  \sigma^{2*}_{t, k} | \sigma^{2*}_{t-1, k} &\sim N \left[\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2) \right]
\end{align}
where $|\phi_w| < 1$, $0 < \phi_z < 1$, and $|\phi_\sigma| < 1$.

@@

\subsection*{Spatial knot locations}
For each day, we update the spatial knot locations, $\bw_1, \ldots, \bw_K$, using a Metropolis Hastings block update.
Because the spatial domain is bounded, we generate candidate knots using the transformed knots $\bw^*_1, \ldots, \bw^*_K$ (see section \ref{s:hier}) and a random walk bivariate Gaussian candidate distribution
\begin{align*}
	{\bw^*_k}^{(c)} \sim \text{N}({\bw^*_k}^{(r - 1)}, s^2 I_2)
\end{align*}
where $\bw*_k^{(r - 1)}$ is the location for the transformed knot at MCMC iteration $r - 1$, $s$ is a tuning parameter, and $I_2$ is an identity matrix.
After candidates have been generated for all $K$ knots, the acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[ Y_t(\bs | \bw_1^{(c)}, \ldots, \bw_K^{(c)}, \ldots)] }{l[ Y_t(\bs | \bw_1^{(r - 1)}, \ldots, \bw_K^{(r - 1)}, \ldots)]} \right\} \times \left\{ \frac{ \prod_{k = 1}^{K}\phi(\bw_k^{(c)})}{ \prod_{k = 1}^{K}\phi(\bw_k^{(r - 1)})} \right\}
\end{align*}
where $l$ is the likelihood given in (\ref{eq:hier}).
The candidate knots are accepted with probability $\min\{R, 1\}$.

\subsection*{Spatial random effects}
If there is temporal dependence amongst the observations, then we update $z_{tk}$, using a Metropolis Hastings update.
First, we generate a candidate using a random walk Gaussian candidate distribution
\begin{align*}
  z_{tk}^{(c)} \sim \text{N}(z_{tk}^{(r - 1)}, s^2)
\end{align*}
where $z_{tk}^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
The acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[Y_t(\bs) | z_{tk}^{(c)}, \ldots] }{ l[Y_t(\bs) | z_{tk}^{(r - 1)}]} \right\} \times \left\{ \frac{ p[ z_{tk}^{(c)} ] }{ p[ z_{tk}^{(r - 1)}]}\right\}.
\end{align*}
The candidate is accepted with probability $\min\{R, 1\}$

>>>>

\subsection{Accounting for temporal dependence} \label{s:temporal}
In a standard block-maxima approach, the temporal dependence is assumed to be negligible because blocks (e.g. yearly maxima) are taken to be large enough to assume independence.
However, when using daily meausrements, the assumption of temporal independence is no longer appropriate.
We account for temporal dependence through an AR(1) time series on the spatial knots, $z_t(\bs)$ terms, and $\sigma^2_t(\bs)$ terms.
We first, transform the spatial knots from $\calD$ to $\calR^2$ as follows.
Let
\begin{align*}
  \bw^*_k = \Phi^{-1}_2\left[ \frac{ (\bw_k - \min(\bs))}{ \text{range}(\bs) } \right].
\end{align*}
where $\Phi_2$ represents a bivariate standard normal density function.
Then $\bw^*_k \in \calR^2$.
We use a copula on the $\sigma^2_t(\bs)$ to ensure that the marginal distributions of $\sigma^2_t(\bs)$ are inverse gamma.
Let
\begin{align*}
  \sigma^{2*}_t(\bs) =\Phi^{-1}\left\{ \text{IG}[\sigma^2_t(\bs)] \right\}
\end{align*}
where $\Phi$ is a univariate standard normal density function, and IG is the distribution function for an IG$(a, b)$ random variable.
Then, when $t > 1$ the time series is modeled as
\begin{align}
  \bw^*_{t, k} | \bw^*_{t-1, k} &\sim N_2\left[\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2) \right] \\
  z_{t, k} | z_{t-1, k} &\sim N \left[\phi_z z_{t-1, k}, \sigma^2_{t,k} (1 - \phi_z^2)\right] \\
  \sigma^{2*}_{t, k} | \sigma^{2*}_{t-1, k} &\sim N \left[\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2) \right]
\end{align}
where $|\phi_w| < 1$, $0 < \phi_z < 1$, and $|\phi_\sigma| < 1$.
All three of the time series are initialized using $N(0, 1)$.

@@

\subsection*{Spatial knot locations}
For each day, we update the spatial knot locations, $\bw_1, \ldots, \bw_K$, using a Metropolis Hastings block update.
Because the spatial domain is bounded, we generate candidate knots using the transformed knots $\bw^*_1, \ldots, \bw^*_K$ (see section \ref{s:temporal}) and a random walk bivariate Gaussian candidate distribution
\begin{align*}
	{\bw^*_k}^{(c)} \sim \text{N}({\bw^*_k}^{(r - 1)}, s^2 I_2)
\end{align*}
where ${\bw^*_k}^{(r - 1)}$ is the location for the transformed knot at MCMC iteration $r - 1$, $s$ is a tuning parameter, and $I_2$ is an identity matrix.
After candidates have been generated for all $K$ knots, the acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[ Y_t(\bs | \bw_1^{(c)}, \ldots, \bw_K^{(c)}, \ldots)] }{l[ Y_t(\bs | \bw_1^{(r - 1)}, \ldots, \bw_K^{(r - 1)}, \ldots)]} \right\} \times \left\{ \frac{ \prod_{k = 1}^{K}\phi(\bw_k^{(c)})}{ \prod_{k = 1}^{K}\phi(\bw_k^{(r - 1)})} \right\} \times \left\{ \frac{ \prod_{k = 1}^{K} p({\bw^*_k}^{(c)})}{ \prod_{k = 1}^{K} p({\bw^*_k}^{(r - 1)})}\right\}
\end{align*}
where $l$ is the likelihood given in (\ref{eq:hier}), and $p(\cdot)$ is the prior either taken from the time series given in (\ref{s:temporal}) or assumed to be uniform over $\calD$.
The candidate knots are accepted with probability $\min\{R, 1\}$.

\subsection*{Spatial random effects}
If there is temporal dependence amongst the observations, then we update $z_{tk}$ using a Metropolis Hastings update.
First, we generate a candidate using a random walk Gaussian candidate distribution
\begin{align*}
  z_{tk}^{(c)} \sim \text{N}(z_{tk}^{(r - 1)}, s^2)
\end{align*}
where $z_{tk}^{(r-1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
The acceptance ratio is
\begin{align*}
  R = \left\{ \frac{ l[Y_t(\bs) | z_{tk}^{(c)}, \ldots] }{ l[Y_t(\bs) | z_{tk}^{(r - 1)}]} \right\} \times \left\{ \frac{ p[ z_{tk}^{(c)} ] }{ p[ z_{tk}^{(r - 1)}]}\right\}.
\end{align*}
The candidate is accepted with probability $\min\{R, 1\}$.

\subsection*{Variance terms}
When there is more than one site in a partition, then we update $\sigma^2_{tk}$ using a Metropolis Hastings update.
First, we generate from one of two candidate distributions.
In the case of temporal dependence, we generate a candidate for ${\sigma^*_{tk}}^2$ using a random walk Gaussian candidate distribution
\begin{align*}
  {\sigma^{*2}_{tk}}^{(c)} \sim \text{N}({\sigma^{*2}_{tk}}^{(r - 1)}, s^2)
\end{align*}
where ${{\sigma^*_{tk}}^2}^{(r - 1)}$ is the value at MCMC iteration $r - 1$, and $s$ is a tuning parameter.
In the case where there is no temporal dependence, we generate a candidate for $\sigma^2_{tk}$ using an IG$(a^*/s, b^*/s)$ candidate distribution where $a^* = n_{tk} / 2 + a$, $b^* = [Y_{tk}^T \Sigma^{-1}_{tk} Y_{tk}] / 2 + b$, $n_{tk}$ is the number of sites in partition $k$ on day $t$, and $Y_{tk}$ and $\Sigma^{-1}_{tk}$ are the observations and precision matrix for partition $k$ on day $t$.
The acceptance ratio is
\begin{align*}
  R = \left\{
    \frac{ l[Y_t(\bs) | {\sigma^2_{tk}}^{(c)}, \ldots] }
         { l[Y_t(\bs) | {\sigma^2_{tk}}^{(r - 1)}]}
    \right\} \times \left\{
    \frac{ l[z_tk | {\sigma^2_{tk}}^{(c)}, \ldots] }
         { l[z_tk | {\sigma^2_{tk}}^{(r - 1)}, \ldots] }
    \right\} \times \left\{
    \frac{ p[ {\sigma^2_{tk}}^{(c)} ] }
         { p[ {\sigma^2_{tk}}^{(r - 1)}] }
    \right\} \times \left\{
    \frac{ c[ {\sigma^2_{tk}}^{(r - 1)}] }
         { c[ {\sigma^2_{tk}}^{(c)}]}
    \right\}
\end{align*}
where $p[\cdot]$ is the prior either taken from the time series given in (\ref{s:temporal}) or assumed to be IG$(a, b)$, and $c[\cdot]$ is the candidate distribution.
The candidate is accepted with probability $\min\{R, 1\}$.
